---
title: "msc_dux4_2025"
author: "Anna Schwager"
date: "`r Sys.Date()`"
output: html_document
---
```{r lib, warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(msigdbr)
library(purrr)
library(patchwork)
library(forcats)
library(tibble)

library(DESeq2)
library(ggplot2)
library(ggrepel)
library(ggrastr)
library(ggvenn)
library(VennDiagram)
library(org.Hs.eg.db)
library(ggplotify)

library(fgsea)
```

# Load data
```{r load, warning=FALSE, message=FALSE}
counts <- read.csv("readcount_MSC_DUX4.csv")
```

# Map ENSEMBL IDs to gene symbols

```{r map, warning=FALSE, message=FALSE}

symbols <- mapIds(org.Hs.eg.db,
                  keys = counts$gene_id,
                  column = "SYMBOL",
                  keytype = "ENSEMBL",
                  multiVals = "first")

counts$gene_symbol <- symbols


num_cols <- which(sapply(counts, is.numeric))

counts <- aggregate(counts[, num_cols],
                    by = list(gene_symbol = counts$gene_symbol),
                    FUN = sum)

rownames(counts) <- counts$gene_symbol
counts$gene_symbol <- NULL

```

# Check library sizes

```{r sizes, warning=FALSE, message=FALSE}
boxplot(log(counts + 1))
```

# Filtering low counts
```{r filt, warning=FALSE, message=FALSE}
dim(counts)

counts_per_gene <- rowSums(counts)
table(counts_per_gene > 0)
counts <- counts[which(counts_per_gene > 0), ]

boxplot(log(counts + 1))
```

# Differential expression analysis 
## DESeq2 
```{r deseq, warning=FALSE, message=FALSE}
cond <- gsub("_[0-9]+$", "", colnames(counts))
cond 

dds <- DESeqDataSetFromMatrix(counts,
                              colData = DataFrame(cond),
                              design = ~ cond) 
colData(dds)
dds <- DESeq(dds)
```

```{r res, warning=FALSE, message=FALSE}
res_DUX4ind_vs_c <- results(dds, contrast = c("cond", "MSC_DUX4_DOX", "ctrl"))
res_DUX4ind_vs_c <- res_DUX4ind_vs_c[complete.cases(res_DUX4ind_vs_c),] 
res_DUX4ind_vs_c <- res_DUX4ind_vs_c[order(res_DUX4ind_vs_c$padj),] 
write.csv(res_DUX4ind_vs_c, "res_DUX4ind_vs_c.csv")
summary(res_DUX4ind_vs_c) 

```
```{r res2, warning=FALSE, message=FALSE}
res_DUX4_vs_c <- results(dds, contrast = c("cond", "MSC_DUX4", "ctrl"))
res_DUX4_vs_c <- res_DUX4_vs_c[complete.cases(res_DUX4_vs_c),] 
res_DUX4_vs_c <- res_DUX4_vs_c[order(res_DUX4_vs_c$padj),] 
write.csv(res_DUX4_vs_c, "res_DUX4_vs_c.csv")
summary(res_DUX4_vs_c) 

```
```{r res3, warning=FALSE, message=FALSE}
res_DUX4ind_vs_DUX4 <- results(dds, contrast = c("cond", "MSC_DUX4_DOX", "MSC_DUX4"))
res_DUX4ind_vs_DUX4 <- res_DUX4ind_vs_DUX4[complete.cases(res_DUX4ind_vs_DUX4),] 
res_DUX4ind_vs_DUX4 <- res_DUX4ind_vs_DUX4[order(res_DUX4ind_vs_DUX4$padj),] 
write.csv(res_DUX4ind_vs_DUX4, "res_DUX4ind_vs_DUX4.csv")
summary(res_DUX4ind_vs_DUX4) 

```

## PCA 
```{r pca, warning=FALSE, message=FALSE}
rld <- rlog(dds) 
plotPCA(rld, intgroup="cond") 

```

## Venn induced vs non-induced
```{r venn, warning=FALSE, message=FALSE}
alpha   <- 0.05           # FDR cutoff
lfc_thr <- 1              # absolute log2FC cutoff


# Significance calls
up1   <- rownames(res_DUX4ind_vs_c)[res_DUX4ind_vs_c$padj < alpha & res_DUX4ind_vs_c$log2FoldChange >=  lfc_thr]
down1 <- rownames(res_DUX4ind_vs_c)[res_DUX4ind_vs_c$padj < alpha & res_DUX4ind_vs_c$log2FoldChange <= -lfc_thr]

up2   <- rownames(res_DUX4ind_vs_DUX4)[res_DUX4ind_vs_DUX4$padj < alpha & res_DUX4ind_vs_DUX4$log2FoldChange >=  lfc_thr]
down2 <- rownames(res_DUX4ind_vs_DUX4)[res_DUX4ind_vs_DUX4$padj < alpha & res_DUX4ind_vs_DUX4$log2FoldChange <= -lfc_thr]

up3 <- rownames(res_DUX4_vs_c)[res_DUX4_vs_c$padj < alpha & res_DUX4_vs_c$log2FoldChange >=  lfc_thr]
down3 <- rownames(res_DUX4_vs_c)[res_DUX4_vs_c$padj < alpha & res_DUX4_vs_c$log2FoldChange <= -lfc_thr]

# Overlaps
up_overlap   <- intersect(up1, up2)
down_overlap <- intersect(down1, down2)

p_up <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = up1,
       `MSC-DUX4 + DOX vs MSC-DUX4 (UP)` = up2),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "grey")
) + ggtitle("Significantly UP genes") 

p_down <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (DOWN)` = down1,
       `MSC-DUX4 + DOX vs MSC-DUX4 (DOWN)` = down2),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "grey")
) + ggtitle("Significantly DOWN genes")

print(p_up)
print(p_down)

```
# DUX4 and PAX7 targets expression
```{r signatures, warning=FALSE, message=FALSE}
dux4_signature_geng <- read.csv("geng_dux4_signature_up.csv")
msig_tft <- msigdbr(species = "Homo sapiens", category = "C3", subcategory = "TFT:GTRD")

pax7_targets <- msig_tft %>%
  filter(gs_name == "PAX7_TARGET_GENES") %>%
  pull(gene_symbol)

```

## Venn diagrams - signatures
```{r signatures2, warning=FALSE, message=FALSE}
venn_dux4_targets <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = up1,
       `DUX4 targets Geng et al` = dux4_signature_geng$Symbol),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "grey")
) + ggtitle("DUX4 targets vs UP genes ") 


venn_dux4_targets_all <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = up1,
       `MSC-DUX4 vs ctrl MSC (UP)` = up3,
       `DUX4 targets Geng et al` = dux4_signature_geng$Symbol),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color = c("#5DADE2", "yellowgreen", "grey")
) + ggtitle("DUX4 targets vs UP genes ") 


venn_pax7_targets <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (DOWN)` = down1,
       `PAX7 targets (MSIG_DB:M30110)` = pax7_targets),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "grey")
) + ggtitle("PAX7 targets vs DOWN genes ") 

venn_pax7_targets_all <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (DOWN)` = down1,
       `MSC-DUX4 vs ctrl MSC (DOWN)` = down3,
       `PAX7 targets (MSIG_DB:M30110)` = pax7_targets),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "yellowgreen", "grey")
) + ggtitle("PAX7 targets vs DOWN genes ") 


p1 <- venn_dux4_targets | venn_pax7_targets 
p2 <- venn_dux4_targets_all | venn_pax7_targets_all

p1
p2

```


```{r save1, eval = FALSE}
ggsave("./plots/venn_dux4.pdf",
       p1)

ggsave("./plots/venn_dux4_all.pdf",
       p2)
```



## Top 12 upregulated DUX4 targets 
```{r dux, warning=FALSE, message=FALSE}
norm_counts <- counts(dds, normalized = TRUE)

res_tbl <- as.data.frame(res_DUX4ind_vs_c) %>%
  tibble::rownames_to_column("symbol")

sample_cond <- setNames(as.character(colData(dds)$cond), colnames(dds))

```


```{r dux2, warning=FALSE, message=FALSE}
top12_dux <- res_tbl %>%
  filter(symbol %in% intersect(up1, dux4_signature_geng$Symbol),
         !is.na(padj), padj < 0.05,
         !is.na(log2FoldChange), log2FoldChange > 0) %>%
  arrange(desc(log2FoldChange), padj) %>%
  slice_head(n = 12) %>%
  pull(symbol)


long_df <- as.data.frame(norm_counts[top12_dux, , drop = FALSE]) %>%
  tibble::rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "sample", values_to = "norm_count") %>%
  mutate(cond = sample_cond[sample]) 

```

```{r dux3, warning=FALSE, message=FALSE}

cols <- c(ctrl = "#E74C3C", MSC_DUX4 = "yellowgreen", MSC_DUX4_DOX = "#5DADE2")

p_box_dux_1 <- ggplot(long_df, aes(x = cond, y = norm_count)) +
  geom_boxplot(aes(fill = cond), outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.2, alpha = 1) +
  scale_fill_manual(values = cols) +
  facet_wrap(~ gene, scales = "free_y", ncol = 4) +
  labs(x = NULL,
       y = "DESeq2 normalized counts",
       title = "Top 12 upregulated DUX4-signature genes") +
  theme_classic(base_size = 12) +
  theme(strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

p_box_dux_1 
```

```{r dux4, warning=FALSE, message=FALSE}

cols <- c(ctrl = "#E74C3C", MSC_DUX4_DOX = "#5DADE2")

p_box_dux_2 <- ggplot(
  long_df %>% filter(cond != "MSC_DUX4"),
  aes(x = cond, y = norm_count)
) +
  geom_boxplot(aes(fill = cond), outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.2, alpha = 1) +
  scale_fill_manual(values = cols) +
  facet_wrap(~ gene, scales = "free_y", ncol = 4) +
  labs(
    x = NULL,
    y = "DESeq2 normalized counts",
    title = "Top 12 upregulated DUX4-signature genes (without MSC_DUX4)"
  ) +
  theme_classic(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )

p_box_dux_2
```

```{r save2, eval =FALSE}
ggsave("./plots/box_dux4.pdf",
       p_box_dux_1,
       width = 20,
      height = 15,
      units = "cm")

ggsave("./plots/box_dux4_main.pdf",
       p_box_dux_2,
       width = 20,
      height = 15,
      units = "cm")
```


## Top 12 downregulated PAX7 targets 
```{r pax, warning=FALSE, message=FALSE}

top12_pax <- res_tbl %>%
  filter(symbol %in% intersect(down1, pax7_targets),
         !is.na(padj), padj < 0.05,
         !is.na(log2FoldChange), log2FoldChange < 0) %>%
  arrange(desc(log2FoldChange), padj) %>%
  slice_head(n = 12) %>%
  pull(symbol)

long_df <- as.data.frame(norm_counts[top12_pax, , drop = FALSE]) %>%
  tibble::rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "sample", values_to = "norm_count") %>%
  mutate(cond = sample_cond[sample]) 

```


```{r pax2, warning=FALSE, message=FALSE}

cols <- c(ctrl = "#E74C3C", MSC_DUX4 = "yellowgreen", MSC_DUX4_DOX = "#5DADE2")

p_box_pax_1 <- ggplot(long_df, aes(x = cond, y = norm_count)) +
  geom_boxplot(aes(fill = cond), outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.2, alpha = 1) +
  scale_fill_manual(values = cols) +
  facet_wrap(~ gene, scales = "free_y", ncol = 4) +
  labs(x = NULL,
       y = "DESeq2 normalized counts",
       title = "Top 12 downregulates PAX7-signature genes") +
  theme_classic(base_size = 12) +
  theme(strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

p_box_pax_1
```

```{r pax3, warning=FALSE, message=FALSE}

cols <- c(ctrl = "#E74C3C", MSC_DUX4_DOX = "#5DADE2")

p_box_pax_2 <- ggplot(
  long_df %>% filter(cond != "MSC_DUX4"),
  aes(x = cond, y = norm_count)
) +
  geom_boxplot(aes(fill = cond), outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.2, alpha = 1) +
  scale_fill_manual(values = cols) +
  facet_wrap(~ gene, scales = "free_y", ncol = 4) +
  labs(
    x = NULL,
    y = "DESeq2 normalized counts",
    title = "Top 12 downregulated PAX7-signature genes (without MSC_DUX4)"
  ) +
  theme_classic(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )

p_box_pax_2
```

```{r save3, eval = FALSE}
ggsave("./plots/box_pax.pdf",
       p_box_pax_1,
      width = 20,
      height = 15,
      units = "cm")

ggsave("./plots/box_pax_main.pdf",
       p_box_pax_2,
      width = 20,
      height = 15,
      units = "cm")
```

# Global GSEA 
Ranking genes
```{r gsea1, warning=FALSE, message=FALSE}
res <- as.data.frame(res_DUX4ind_vs_c)
res$gene_id <- rownames(res)

res <- res %>% arrange(desc(log2FoldChange))

ranks <- res$log2FoldChange
names(ranks) <- res$gene_id
```

## Hallmark gene sets
```{r gsea2, warning=FALSE, message=FALSE}
msig_h <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  distinct()

hallmark <- msig_h %>%
  group_by(gs_name) %>%
  summarise(genes = list(unique(gene_symbol)), .groups = "drop") %>%
  { setNames(.$genes, .$gs_name) }

```

```{r gsea3, warning=FALSE, message=FALSE}
fg_all <- fgsea(
  pathways = hallmark,
  stats    = ranks,
  minSize  = 5,
  maxSize  = 2000,
) %>% arrange(padj)

head(fg_all[, c("pathway","size","NES","pval","padj")], 10)


fg_all_out <- fg_all %>%
  as.data.frame() %>%
  mutate(leadingEdge = vapply(leadingEdge, function(x) paste(x, collapse = ";"), character(1))) %>%
  arrange(padj)

#write.csv(fg_all_out, "GSEA_Hallmark_results.csv", row.names = FALSE)

```
```{r gsea4, warning=FALSE, message=FALSE}

# all significant pathways
fg_plot <- fg_all %>%
  mutate(pathway_clean = gsub("^HALLMARK_", "", pathway),
         pathway_clean = gsub("_", " ", pathway_clean)) %>%
  arrange(padj) %>%
  slice_head(n = 15) %>%
  mutate(pathway_clean = fct_reorder(pathway_clean, NES))

p4 <- ggplot(fg_plot, aes(x = pathway_clean, y = NES, fill = padj)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(
    low = "#E74C3C",
    high = "#5DADE2", 
    trans = "log10",
    name = "Adjusted p-value"
  ) +
  labs(
    x = NULL,
    y = "Normalized Enrichment Score (NES)",
    title = "Significantly enriched Hallmark pathways"
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 10),
    plot.title = element_text(size = 13)
  )

p4
```
```{r save4, eval=FALSE}
ggsave("./plots/gsea_all.pdf",
       p4,
      width = 20,
      height = 10,
      units = "cm")

```

# Programs expression

```{r prog1, warning=FALSE, message=FALSE}
msig_hallmark <- msigdbr(species = "Homo sapiens", category = "H")
msig_bp <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP")
msig_curated <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CGP")


myo_genes <- msig_hallmark %>%
  filter(gs_name == "HALLMARK_MYOGENESIS") %>%
  pull(gene_symbol)

adip_genes <- msig_hallmark %>%
  filter(gs_name == "HALLMARK_ADIPOGENESIS") %>%
  pull(gene_symbol)

angio_genes <- msig_hallmark %>%
  filter(gs_name == "HALLMARK_ANGIOGENESIS") %>%
  pull(gene_symbol)

osteo_genes <- msig_bp %>%
  filter(gs_name == "GOBP_OSTEOBLAST_DIFFERENTIATION") %>%
  pull(gene_symbol)

chondro_genes <- msig_bp %>%
  filter(gs_name == "GOBP_CHONDROCYTE_DIFFERENTIATION") %>%
  pull(gene_symbol)

```

## Lineage programs
### Up
```{r up, warning=FALSE, message=FALSE}
p_up_myo <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = up1,
       `HALLMARK_MYOGENESIS` = myo_genes),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "#E74C3C")
) + ggtitle("UP genes vs Myogenesis genes") 

p_up_adip <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = up1,
       `HALLMARK_ADIPOGENESIS` = adip_genes),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "#FFC125")
) + ggtitle("UP genes vs Adipogenesis genes") 

p_up_osteo <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = up1,
       `GOBP_OSTEOBLAST_DIFFERENTIATION` = osteo_genes),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "#8B795E")
) + ggtitle("UP genes vs Osteogenesis genes")

p_up_chondro <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = up1,
       `GOBP_CHONDROCYTE_DIFFERENTIATION` = chondro_genes),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "plum")
) + ggtitle("UP genes vs Chondrogenesis genes")


p_prog_up <- p_up_osteo / p_up_adip / p_up_chondro / p_up_myo 
p_prog_up

```

```{r save5, eval=FALSE}
ggsave("./plots/p_prog_upl.pdf",
      p_prog_up,
      width = 20,
      height = 15,
      units = "cm")

```

### Down
```{r down, warning=FALSE, message=FALSE}
p_down_myo <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = down1,
       `HALLMARK_MYOGENESIS` = myo_genes),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "#E74C3C")
) + ggtitle("DOWN genes vs Myogenesis genes") 

p_down_adip <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = down1,
       `HALLMARK_ADIPOGENESIS` = adip_genes),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "#FFC125")
) + ggtitle("DOWN genes vs Adipogenesis genes") 

p_down_osteo <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = down1,
       `GOBP_BONE_CELL_DEVELOPMENT` = osteo_genes),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "#8B795E")
) + ggtitle("DOWN genes vs Osteogenesis genes")


p_down_chondro <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = down1,
       `GOBP_CHONDROCYTE_DEVELOPMENT` = chondro_genes),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "plum")
) + ggtitle("DOWN genes vs Chondrogenesis genes")

p_prog_down <- p_down_osteo / p_down_adip / p_down_chondro / p_down_myo 
p_prog_down

```

```{r save6, eval=FALSE}
ggsave("./plots/p_prog_down.pdf",
      p_prog_down,
      width = 20,
      height = 15,
      units = "cm")

```



## Volcano with myogenesis genes highlighted

```{r volc1, warning=FALSE, message=FALSE}

bp <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:BP") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  distinct()

bp <- bp %>%
  group_by(gs_name) %>%
  summarise(genes = list(unique(gene_symbol)), .groups = "drop") %>%
  { setNames(.$genes, .$gs_name) }

myo_der_genes <- intersect(up1, myo_genes)
GOBP_NEGATIVE_REGULATION_OF_SMOOTH_MUSCLE_CELL_DIFFERENTIATION <- intersect(down1, bp$GOBP_NEGATIVE_REGULATION_OF_SMOOTH_MUSCLE_CELL_DIFFERENTIATION)

```


```{r volc2, warning=FALSE, message=FALSE}

res_tbl <- as.data.frame(res_DUX4ind_vs_c) |>
  tibble::rownames_to_column("gene") |>
  mutate(group = case_when(
    gene %in% myo_der_genes        ~ "HALLMARK_MYOGENESIS",
    gene %in% GOBP_NEGATIVE_REGULATION_OF_SMOOTH_MUSCLE_CELL_DIFFERENTIATION      ~ "GOBP_NEG_REG_OF_SM_MUS_C_DIFF",
    TRUE                          ~ "Other"
  ))

# avoid Inf in -log10
res_tbl$padj[res_tbl$padj == 0] <- .Machine$double.xmin

res_bg <- res_tbl |> filter(group == "Other")
res_fg <- res_tbl |> filter(group != "Other")

cols <- c(HALLMARK_MYOGENESIS = "#E74C3C",
          GOBP_NEG_REG_OF_SM_MUS_C_DIFF = "#5DADE2")

```


```{r volc3, warning=FALSE, message=FALSE}

p_volcano <- ggplot() +
  # 1) grey background points FIRST
  geom_point(data = res_bg,
             aes(x = log2FoldChange, y = -log10(padj)),
             color = "grey75", alpha = 0.6, size = 1.5) +
  # 2) colored groups ON TOP
  geom_point(data = res_fg,
             aes(x = log2FoldChange, y = -log10(padj), color = group),
             alpha = 0.95, size = 2.2) +
  scale_color_manual(values = cols, breaks = names(cols), name = NULL) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black", linewidth = 0.5) +
  labs(title = "MSC-DUX4 induced vs Control MSC",
       x = "log2 Fold Change",
       y = expression(-log[10](adjusted~p~value))) +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
  # labels only for colored groups
  geom_text_repel(data = res_fg,
                  aes(x = log2FoldChange, y = -log10(padj), label = gene, color = group),
                  size = 3, box.padding = 0.3, point.padding = 0.2,
                  segment.color = "grey70", max.overlaps = 20, show.legend = FALSE)

p_volcano


p_volcano_rast <- ggplot() +
  geom_point_rast(data = res_bg,
                  aes(x = log2FoldChange, y = -log10(padj)),
                  color = "grey75", alpha = 0.6, size = 1.5,
                  raster.dpi = 300) +
  geom_point_rast(data = res_fg,
                  aes(x = log2FoldChange, y = -log10(padj), color = group),
                  alpha = 0.95, size = 2.2,
                  raster.dpi = 300) +
  scale_color_manual(values = cols, breaks = names(cols), name = NULL) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black", linewidth = 0.5) +
  labs(title = "MSC-DUX4 induced vs Control MSC",
       x = "log2 Fold Change",
       y = expression(-log[10](adjusted~p~value))) +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
  geom_text_repel(data = res_fg,
                  aes(x = log2FoldChange, y = -log10(padj), label = gene, color = group),
                  size = 3, box.padding = 0.3, point.padding = 0.2,
                  segment.color = "grey70", max.overlaps = 20, show.legend = FALSE)
```
```{r save7, eval = FALSE}
ggsave("./plots/myo_volcano.pdf",
       p_volcano,
      width = 18,
      height = 10,
      units = "cm")

ggsave("./plots/mvolcano_rasterized.pdf",
      p_volcano_rast,
      width = 20,
      height = 12,
      units = "cm")

```


## ECM genes expression 

```{r ecm1, warning=FALSE, message=FALSE}
core_matrisome <- msig_curated %>%
  filter(gs_name == "NABA_CORE_MATRISOME") %>%
  pull(gene_symbol)

col_genes <- msig_curated %>%
  filter(gs_name == "NABA_COLLAGENS") %>%
  pull(gene_symbol)

gly_genes <- msig_curated %>%
  filter(gs_name == "NABA_ECM_GLYCOPROTEINS") %>%
  pull(gene_symbol)

ecm_reg_genes <- msig_curated %>%
  filter(gs_name == "NABA_ECM_REGULATORS") %>%
  pull(gene_symbol)


```

```{r ecm2, warning=FALSE, message=FALSE}
p_ecm <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = up1,
       `MSC-DUX4 + DOX vs ctrl MSC (DOWN)` = down1,
       `core_matrisome` = core_matrisome),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "#5DADE2","orchid")
) + ggtitle("Diff genes vs Core matrisome genes") 

p_ecm_reg <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = up1,
       `MSC-DUX4 + DOX vs ctrl MSC (DOWN)` = down1,
       `NABA_ECM_REGULATORS` = ecm_reg_genes),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2", "#5DADE2","orchid")
) + ggtitle("Diff genes vs ECM regulators genes") 

p_col <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = up1,
       `MSC-DUX4 + DOX vs ctrl MSC (DOWN)` = down1,
       `NABA_COLLAGENS` = col_genes),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2","#5DADE2","orchid")
) + ggtitle("Diff genes vs Collagen genes") 

p_gly <- ggvenn(
  list(`MSC-DUX4 + DOX vs ctrl MSC (UP)` = up1,
       `MSC-DUX4 + DOX vs ctrl MSC (DOWN)` = down1,
       `NABA_NABA_ECM_GLYCOPROTEINS` = gly_genes),
  show_percentage = FALSE,
  set_name_size = 3.5,
  fill_color   = c("#5DADE2","#5DADE2","orchid")
) + ggtitle("Diff genes vs Collagen genes") 

ecm <- p_ecm | p_col | p_gly | p_ecm_reg
ecm
```
```{r save8, eval = FALSE}
ggsave("./plots/ecm_venn.pdf",
      ecm,
      width = 40,
      height = 15,
      units = "cm")
```

## Volcano with ECM genes

```{r volc_ecm, warning=FALSE, message=FALSE}
col_genes_der <- intersect(c(up1,down1), col_genes)
gly_genes_der <- intersect(c(up1,down1), gly_genes)

```


```{r volc_ecm2, warning=FALSE, message=FALSE}

res_tbl <- as.data.frame(res_DUX4ind_vs_c) |>
  tibble::rownames_to_column("gene") |>
  mutate(group = case_when(
    gene %in% col_genes_der        ~ "NABA_COLLAGENS",
    gene %in% gly_genes_der      ~ "NABA_ECM_GLYCOPROTEINS",
    TRUE                          ~ "Other"
  ))

# avoid Inf in -log10
res_tbl$padj[res_tbl$padj == 0] <- .Machine$double.xmin

res_bg <- res_tbl |> filter(group == "Other")
res_fg <- res_tbl |> filter(group != "Other")


label_genes <- bind_rows(
  res_fg |> filter(group == "NABA_COLLAGENS"),
  res_fg |> 
    filter(group == "NABA_ECM_GLYCOPROTEINS") |> 
    arrange(desc(abs(log2FoldChange))) |> 
    slice_head(n = 10)
)

cols <- c(NABA_COLLAGENS = "orchid4",
          NABA_ECM_GLYCOPROTEINS = "lightcoral")

```


```{r volc_ecm3, warning=FALSE, message=FALSE}

p_volcano_ecm <- ggplot() +
  geom_point(data = res_bg,
             aes(x = log2FoldChange, y = -log10(padj)),
             color = "grey75", alpha = 0.6, size = 1.5) +
  geom_point(data = res_fg,
             aes(x = log2FoldChange, y = -log10(padj), color = group),
             alpha = 0.95, size = 2.2) +
  scale_color_manual(values = cols, breaks = names(cols), name = NULL) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black", linewidth = 0.5) +
  labs(title = "MSC-DUX4 induced vs Control MSC",
       x = "log2 Fold Change",
       y = expression(-log[10](adjusted~p~value))) +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
  geom_text_repel(
    data = label_genes,
    aes(x = log2FoldChange, y = -log10(padj), label = gene, color = group),
    size = 3, box.padding = 0.3, point.padding = 0.2,
    segment.color = "grey70", max.overlaps = 30, show.legend = FALSE
  )

p_volcano_ecm


p_volcano_ecm_rast <- ggplot() +
  geom_point_rast(data = res_bg,
             aes(x = log2FoldChange, y = -log10(padj)),
             color = "grey75", alpha = 0.6, size = 1.5) +
  geom_point_rast(data = res_fg,
             aes(x = log2FoldChange, y = -log10(padj), color = group),
             alpha = 0.95, size = 2.2) +
  scale_color_manual(values = cols, breaks = names(cols), name = NULL) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black", linewidth = 0.5) +
  labs(title = "MSC-DUX4 induced vs Control MSC",
       x = "log2 Fold Change",
       y = expression(-log[10](adjusted~p~value))) +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
  geom_text_repel(
    data = label_genes,
    aes(x = log2FoldChange, y = -log10(padj), label = gene, color = group),
    size = 3, box.padding = 0.3, point.padding = 0.2,
    segment.color = "grey70", max.overlaps = 30, show.legend = FALSE
  )

```
```{r save 9, eval = FALSE}

ggsave("./plots/ecm_volcano_rasterized.pdf",
      p_volcano_ecm_rast,
      width = 20,
      height = 12,
      units = "cm")

```


